A Standard for the Scholarly Citation of Archaeological Data

### 2017-09-12 analysis/paper/paper.Rmd
arrange plots for email requests, article survey, cowplot is great for this

@@

Defining what data are is an ongoing research problem for information scientists (Leonelli 2016; Borgman 2012). Table X summarises some of the definitions debated in the philosophy of science literature. A more concrete starting point is this definition from the National Academies of Sciences:

@@

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different requirements for access, archiving, and so on. For many archaeologists, data are the contents of the files we use to generate visualisations, summary tables, and qualitative statements that appear in, and support the claims we make in, our scholarly publications and reports. The categories shown in Table X are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: “What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models.” This is a highly pragmatic approach to defining data, but is also consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependant on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). It implies that the answer to the question of what counts as data is mostly emergent from our professional interactions, such as the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email.  Our citation standard imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion. Table X summarises some desirable minimum qualities for archaeological data that optimise its discovery and reuse.

@@

While public and community benefits are well-rehearsed and widely agreed on (Table X), they offer no, or only weak, tangible incentives to the individual researcher (Rowhani-Farid et al. 2017). Previous work has identified citation advantages as a tangible benefit to individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table X for a summary). Increased publication productivity also appeared to be associated with data sharing behaviours. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only 5 for projects without archived data.

@@

Sharing of data is a quality that is fundamental to the definition of science (cf. Merton’s (1942) communalism norm, where intellectual property rights are normally given up in exchange for recognition and esteem) and to data both as an abstract quality (cf. the definitions in Table X that mention data as ‘portable’, ‘stable across different domains’) and as a concrete detail to be managed with policy. For example, for over a decade the National Science Foundation (2016) has had a data sharing policy that states “Investigators are expected to share with other researchers, at no more than incremental cost and within a reasonable time, the primary data, samples, physical collections and other supporting materials created or gathered in the course of work under NSF grants. Grantees are expected to encourage and facilitate such sharing.” Surveys of researchers across different domains repeatedly show that people value the ideal of data sharing (Kratz and Strasser 2015). However, data sharing is not widespread in many disciplines, raising questions about why researchers do or do not share their data. Three recent studies have attempted to answer these questions, in the search for strategies to encourage sharing.

Stodden (2010) surveyed registrants at a top machine learning conference (the Neural Information Processing Systems (NIPS) conference) to identify the factors that affect researchers’ decisions to reveal data (Table X). Stodden found that communitarian factors influence scientists to share their data, but private incentives determine when scientists choose not to share their code and data. In the top three reasons not to share, we see a lack of incentives to invest time in data sharing (however, there are many other time-consuming tasks in research, and they get done because clear reward structures are in place); a lack of faith in data citation (which we hope this paper will help with); and a lack of an efficient copyright process. On this last point, Table X shows some of the current norms on this, based on Stodden’s (2009a) detailed legal discussion. See also Stodden’s (2009b) ‘Reproducible Research Standard’, summarised in Figure X for a convenient and robust copyright schema for typical research products.

@@

Even though there is a wide spectrum of data sharing behaviours, from intensive to restrictive, overall rates of data sharing are very low in many fields. Wicherts et al. (2006) emailed authors of 141 empirical articles recently published by the American Psychological Association (APA) journals to request a copy of the original data reported in the articles. They received 64 of the 249 datasets they requested, resulting in a data sharing rate of about 26%, despite all the authors of the articles in their study being bound to the APA’s Ethical Principles, which state authors must ‘not withhold the data on which their conclusions are based from other competent professionals who seek to verify the substantive claims’ (American Psychological Association, 2001, p. 396). Valpaemel et al. (2015) similarly requested data from 394 papers, published in all issues of four APA journals in 2012 and found that 38% of authors responded with their data. They concluded that although the data sharing rate  increased as compared to the study by Wicherts et al., their findings are ‘worrisome’ and indicate a ‘poor availability of data’.

Savage and Vickers (2009) requested data from authors of ten papers published in PLOS journals, and received only one reply. Although this is a very small sample, it is notable because PLOS journals require authors to share data as a condition of publication: “PLOS journals require authors to make all data underlying the findings described in their manuscript fully available without restriction, with rare exception.” (PLoS Medicine Editorial and Publishing Policies nd), and yet only one out of the ten authors actually complied with this requirement. Alsheikh-Ali et al. (2011) inspected 500 articles published in high-impact science journals in 2009 and found that only 47 papers (9%) deposited full primary raw data online. Similarly, Womak (2015) inspected 4,370 articles published in high-impact science journals in 2014 and found that 13% of articles with original data published make the data available to others.

@@

Here we present the results of three small pilot studies aimed at understanding how archaeologists share their data. First, we emailed a sample of authors of research publications to request access to the data behind their publication. Second, we surveyed a sample of journal articles to see what data was available. Third, we analysed information about archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data).

@@

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  eval = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

```{r import-email-data, eval = TRUE}
library(tidyverse)
email_survey <- readr::read_csv("../data/raw_data/email_survey.csv")

@@

Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the one PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014): "Publication is conditional upon the agreement of the authors to make freely available any materials and information described in their publication that may be reasonably requested by others for the purpose of academic, non-commercial research." Chambers' (2017) examined 50 PLOS One papers on brain images published in November 2014 - May 2015 and found that only 38% of these papers archived their data in an open repository. Similarly, Nuijten et al. (2017) found that of 462 PLOS One articles that promised data availability, 29% of these did not have any available data. Unfulfilled promised to share data are not unique to PLOS One (Kidwell et al. 2016), but these studies highlight the need to resolve ambiguity among authors about what counts as data, and for editors and peer reviwers to more critically assess whether an author has been made their data openly available.

@@

```{r date-and-sharing, fig.cap="There is no obvious trend in data sharing becavious and date of publication", eval = TRUE}
library(viridis)
years <- seq(min(email_survey$'publication year', na.rm = TRUE),
    max(email_survey$'publication year', na.rm = TRUE), 1)
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
  group_by(`publication year`, `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`publication year`)) %>% 
  mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  ggplot(aes(`publication year`, n)) +
    geom_col(aes(fill = `content recieved date`),
             # need devtools::install_github("tidyverse/ggplot2") for this
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = guide_legend(title = "Data shared?")) +
    scale_x_continuous(breaks = years,
                       labels = years) +
    theme_bw()

@@

```{r student, fig.cap="No relationship between student status and data sharing behaviour", eval = TRUE}
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  `author is graduate student at time of publication?`, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`author is graduate student at time of publication?`)) %>% 
  ggplot(aes(`author is graduate student at time of publication?`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = guide_legend(title = "Data shared?")) +
    theme_bw()
```

```{r author-country, eval = TRUE}
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  `author home`, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`author home`)) %>% 
  ggplot(aes(`author home`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = guide_legend(title = "Data shared?")) +
    theme_bw()
```

```{r journals, fig.cap="No relationship between sharing behaviour and journal.", eval = TRUE}
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  journal, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(journal)) %>% 
  ggplot(aes(journal, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = guide_legend(title = "Data shared?")) +
    theme_bw()
```

@@

```{r eval = TRUE}
jas_article_survey <- readr::read_csv("../data/raw_data/jas_article_survey.csv")

@@

```{r eval=TRUE, fig.cap="No relationship between data file type and data sharing location"}
# correlation of data type and location of sharing
file_type_location <- 
jas_article_survey %>% 
  mutate(file = strsplit(as.character(`What file format?`), ",")) %>% 
  unnest(file) %>% 
  filter(file != "na" & `Where is the data?` != "na") %>% 
  select(file,  `Where is the data?`) %>% 
  mutate(location = case_when(
    grepl("sup", tolower(`Where is the data?`)) ~ "Supplementary",
    grepl("tab", tolower(`Where is the data?`)) ~ "Table",
    grepl("www|dryad|osf", tolower(`Where is the data?`)) ~ "Repository",
  )) %>% 
  mutate(file = case_when(
    grepl("DOCX|doc", file) ~ "DOCX",
    grepl("XLSX| XLSX", file) ~ "XLSX",
    TRUE ~ file
    ))

file_type_location %>% 
  group_by(file, location) %>% 
  tally() %>% 
  ggplot(aes(file, location)) + 
  geom_point(aes(size = n),
             colour = "grey80") +
  theme_bw() + 
  xlab("") + 
  ylab("") +
  scale_size_continuous(range=c(5,25)) + 
  geom_text(aes(label = n)) +
  theme(legend.position = "none") 
```

```{r eval=TRUE, fig.cap="Plot of the relationship between data type and location where it is shared."}
# correlation of data type and location of sharing
data_type_location <- 
jas_article_survey %>% 
  filter(`data type?` != "na" & `Where is the data?` != "na") %>% 
  select(`data type?`, `Where is the data?`) %>% 
  mutate(data_type = case_when(
    grepl("comp", tolower(`data type?`)) ~ "Compositional",
    grepl("c14|osl", tolower(`data type?`)) ~ "Dating",
    grepl("dna", tolower(`data type?`)) ~ "DNA",
    TRUE ~ `data type?`
  )) %>% 
   mutate(location = case_when(
    grepl("sup", tolower(`Where is the data?`)) ~ "Supplementary",
    grepl("tab", tolower(`Where is the data?`)) ~ "Table",
    grepl("www|dryad|osf", tolower(`Where is the data?`)) ~ "Repository"
  ))

data_type_location %>% 
  group_by(data_type, location) %>% 
  tally() %>% 
  filter(n >= 2) %>% 
  ggplot(aes(data_type, location)) + 
  geom_point(aes(size = n),
             colour = "grey80") +
  theme_bw() + 
  xlab("") + 
  ylab("") +
  scale_size_continuous(range=c(5,25)) + 
  geom_text(aes(label = n)) +
  theme(legend.position = "none") 
```

@@

A full treatment of the ethical issues of data sharing is beyond the scope of this paper, but we can highlight some points specifically relevant to the question of when archaeological data. Much of the discussion on ethics and data sharing generally relates to appropriate attribution (assigning credit for the data) in data reuse (Duke et al. 2013). The default position for archaeological data, like data from many other sciences, is that they are part of the global human patrimony, and should be accessible to other scholars and the public (Zimmerman et al. 2003; Vitelli & Colwell-Chanthaphonh 2006). However, archaeologists, like other researchers working with indigenous communities and governments, generate research data that is intellectual property with many stakeholders. With diverse stakeholders, archaeological data can become entangled in economics, nationalism, cultural politics, and identity (Nicholas and Wylie 2009). Often, one of these stakeholders is an indigenous community, and archaeologists may find differences between Western and indigenous conceptions of what data are, how they may be generated, used, shared, and 'owned' (Harding et al. 2012).

One of these differences is that Western researchers often have individual autonomy in sharing data, and once data are shared, it is free for all to use. In contrast, within indigenous communities, data are part of the group's identity and property, akin to tangible hertiage (Nicholas 2014), with certain members entrusted to keep that data on behalf of the group, and sometimes it may be inappropriate for these people to share the data with other members of the group, or beyond the group (Tsosie 2007). Differences between Western and indigenous concepts or data are not only cultural - substantial inequalities exist in access to resources for converting archaeological data into economic and political benefit (e.g. graduate degrees, employment, publications, grants, promotions). On one hand, making archaeological data openly available benefits indigenous communities, because it removes some of the financial barriers to accessing that data (e.g. subscription fees to journals). On the other hand, this may be little more than a useless courtesy (Hymes 1972), because simply making data openly available does not give indigenous people access to the contexts where the data can be used to generate economic and political benefits. Thus, open data has different values for Western and indigenous communities, and if archaeologists want to make a compelling case to their indigenous partners for data sharing, they should also make a committement to reducing some of the inequalities in making use of the data, for example, by providing some training in data analysis, or adapting the data into a context that is valued by the indigenous community.

In short, there is more to the ethics of data sharing than avoiding risk of damage. Redacting the locations of archaeological sites to prevent looting as an obvious case of preparing data for sharing to reduce the risk of damage to the sites. However, because of cultural, economic and political differences between Western archaeologists and the communities they work with, an archaeologist cannot consider their own risk assessment to be comprehensive (Nicholas and Wylie 2012). It is crucial for archaeologists to discuss data sharing plans during negotiatations with representatives authorized by the indigenous peoples whose cultural heritage is the subject of investigation. This negotiation process is important because only indigenous peoples themselves can identify potential adverse outcomes to data sharing, and they can do this only if they understand the proposed research and anticipated results. The concrete outcome of this process is often a Memorandum of Understanding between the indigenous community and the archaeologist's employer that documents the ourcome of negotiations about data sharing. If the archaeologist's indigenous partners have reasons not to share data in public, then the archaeologist should respect that position, and note in publication that the data are restricted at the request of the indigenous collaborators.

@@

Hart EM, Barmby P, LeBauer D, Michonneau F, Mount S, et al. (2016) Ten Simple Rules for Digital Data Storage. PLOS Computational Biology 12(10): e1005097. doi: 10.1371/journal.pcbi.1005097

>>>>

```{r, setup2, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  eval = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

Defining what data are is an ongoing research problem for information scientists (Leonelli 2016; Borgman 2012). Table \@ref(tab:summary-of-defs-of-data) summarises some of the definitions debated in the philosophy of science literature. A more concrete starting point is this definition from the National Academies of Sciences:

@@

```{r summary-of-defs-of-data}
tbl <- readr::read_csv("../figures/tbl_definitions_of_data.csv")
knitr::kable(tbl, caption = "A selection of defintions of research data")
```

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different requirements for access, archiving, and so on. For many archaeologists, data are the contents of the files we use to generate visualisations, summary tables, and qualitative statements that appear in, and support the claims we make in, our scholarly publications and reports. The categories shown in Table \@ref(tab:types-of-data) are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: "What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models." This is a highly pragmatic approach to defining data, but is also consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependant on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). It implies that the answer to the question of what counts as data is mostly emergent from our professional interactions, such as the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email.  Our citation standard imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion. Table \@ref(tab:desirable) summarises some desirable minimum qualities for archaeological data that optimise its discovery and reuse.

```{r types-of-data}
tbl <- readr::read_csv("../figures/tbl_types_of_data_nsb.csv")
knitr::kable(tbl, caption = "A selection of types of data encountereed by archaeologists.")
```

@@

While public and community benefits and costs are well-rehearsed and widely agreed on (Table \@ref(tab:costsbenefits)), they offer no, or only weak, tangible incentives to the individual researcher (Rowhani-Farid et al. 2017). Previous work has identified citation advantages as a tangible benefit to individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table \@ref(tab:indivbenefits) for a summary). Increased publication productivity also appeared to be associated with data sharing behaviours. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only five for projects without archived data.

```{r costsbenefits}
tbl <- readr::read_csv("../figures/tbl_reasons_to_share_data.csv")
knitr::kable(tbl, caption = "Summary of public and community benefits and costs of data sharing")
```

```{r indivbenefits}
tbl <- readr::read_csv("../figures/tbl_evidence_of_sharing_benefits.csv")
knitr::kable(tbl, caption = "Summary of individual benefits to data sharing")
```

@@

Sharing of data is a quality that is fundamental to the definition of science (cf. Merton’s (1942) communalism norm, where intellectual property rights are normally given up in exchange for recognition and esteem) and to data both as an abstract quality (cf. the definitions in Table \@ref(tab:summary-of-defs-of-data) that mention data as ‘portable’, ‘stable across different domains’) and as a concrete detail to be managed with policy. For example, for over a decade the National Science Foundation (2016) has had a data sharing policy that states "Investigators are expected to share with other researchers, at no more than incremental cost and within a reasonable time, the primary data, samples, physical collections and other supporting materials created or gathered in the course of work under NSF grants. Grantees are expected to encourage and facilitate such sharing." Surveys of researchers across different domains repeatedly show that people value the ideal of data sharing (Kratz and Strasser 2015). However, data sharing is not widespread in many disciplines, raising questions about why researchers do or do not share their data. Three recent studies have attempted to answer these questions, in the search for strategies to encourage sharing.

Stodden (2010) surveyed registrants at a top machine learning conference (the Neural Information Processing Systems (NIPS) conference) to identify the factors that affect researchers’ decisions to reveal data (Table \@ref(tab:costsbenefits)). Stodden found that communitarian factors influence scientists to share their data, but private incentives determine when scientists choose not to share their code and data. In the top three reasons not to share, we see a lack of incentives to invest time in data sharing (however, there are many other time-consuming tasks in research, and they get done because clear reward structures are in place); a lack of faith in data citation (which we hope this paper will help with); and a lack of an efficient copyright process. On this last point, Table X shows some of the current norms on this, based on Stodden’s (2009a) detailed legal discussion. See also Stodden’s (2009b) ‘Reproducible Research Standard’, summarised in Figure \@(ref:stodden-rrs) for a convenient and robust copyright schema for typical research products.

```{r stodden-rrs, fig.cap="Summary of Stodden’s (2009b) 'Reproducible Research Standard'"}
knitr::include_graphics("../figures/fig1_rrs.png")
```

@@

Even though there is a wide spectrum of data sharing behaviours, from intensive to restrictive, overall rates of data sharing are very low in many fields. Wicherts et al. (2006) emailed authors of 141 empirical articles recently published by the American Psychological Association (APA) journals to request a copy of the original data reported in the articles. They received 64 of the 249 datasets they requested, resulting in a data sharing rate of about 26%, despite all the authors of the articles in their study being bound to the APA's Ethical Principles, which state authors must ‘not withhold the data on which their conclusions are based from other competent professionals who seek to verify the substantive claims’ (American Psychological Association, 2001, p. 396). Valpaemel et al. (2015) similarly requested data from 394 papers, published in all issues of four APA journals in 2012 and found that 38% of authors responded with their data. They concluded that although the data sharing rate increased as compared to the study by Wicherts et al., their findings are ‘worrisome’ and indicate a 'poor availability of data'.

Savage and Vickers (2009) requested data from authors of ten papers published in PLOS journals, and received only one reply. Although this is a very small sample, it is notable because PLOS journals require authors to share data as a condition of publication: "PLOS journals require authors to make all data underlying the findings described in their manuscript fully available without restriction, with rare exception." (PLoS Medicine Editorial and Publishing Policies nd), and yet only one out of the ten authors actually complied with this requirement. Alsheikh-Ali et al. (2011) inspected 500 articles published in high-impact science journals in 2009 and found that only 47 papers (9%) deposited full primary raw data online. Similarly, Womak (2015) inspected 4,370 articles published in high-impact science journals in 2014 and found that 13% of articles with original data published make the data available to others.

@@

Here we present the results of three small pilot studies aimed at understanding how archaeologists share their data. First, we emailed a sample of authors of research publications to request access to the data behind their publication. Second, we surveyed a sample of journal articles to see what data was available. Third, we analysed information about archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The small sample sizes of the first two of these pilot studies limit how generalisable the results are, but as the first of their kind for archaeology, they are valuable for proposing hypotheses for larger-scale investigations.

@@

```{r import-email-data}
library(tidyverse)
email_survey <- readr::read_csv("../data/raw_data/email_survey.csv")

@@

Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the one PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014): "Publication is conditional upon the agreement of the authors to make freely available any materials and information described in their publication that may be reasonably requested by others for the purpose of academic, non-commercial research." Chambers' (2017) examined 50 PLOS One papers on brain images published in November 2014 - May 2015 and found that only 38% of these papers archived their data in an open repository. Similarly, Nuijten et al. (2017) found that of 462 PLOS One articles that promised data availability, 29% of these did not have any available data. Unfulfilled promises to share data are not unique to PLOS One (Kidwell et al. 2016), but these studies highlight the need to resolve ambiguity among authors about what counts as data, and for editors and peer reviwers to more critically assess whether an author has been made their data openly available.

@@

```{r date-and-sharing, fig.cap="There is no obvious trend in data sharing becavious and date of publication"}
library(viridis)
years <- seq(min(email_survey$'publication year', na.rm = TRUE),
    max(email_survey$'publication year', na.rm = TRUE), 1)

sharing_by_year <- 
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
  group_by(`publication year`, `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`publication year`)) %>% 
  mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  ggplot(aes(`publication year`, n)) +
    geom_col(aes(fill = `content recieved date`),
             # need devtools::install_github("tidyverse/ggplot2") for this
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = FALSE) +
    scale_x_continuous(breaks = years,
                       labels = years) +
  xlab("Publication year") +
    theme_bw() +
  coord_fixed(ratio = 0.5)

@@

```{r student, fig.cap="No relationship between student status and data sharing behaviour"}
sharing_by_student <- 
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  `author is graduate student at time of publication?`, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`author is graduate student at time of publication?`)) %>% 
  ggplot(aes(`author is graduate student at time of publication?`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = FALSE) +
    xlab("Author was graduate student at time of publication?") +
    theme_bw() +
  coord_fixed(ratio = 0.05)
```

```{r author-country}
sharing_by_country <- 
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  `author home`, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`author home`)) %>% 
  ggplot(aes(`author home`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = FALSE) +
    theme_bw() +
  xlab("Author location at time of publication") +
  coord_fixed(ratio = 0.25)
```

```{r journals, fig.cap="No relationship between sharing behaviour and journal."}
sharing_by_journal <- 
  email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  journal, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(journal)) %>% 
  ggplot(aes(journal, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = guide_legend(title = "Data shared?")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = c(0.8, 0.7)) +
  xlab("Journal name") +
  coord_fixed(ratio = 0.25)
```

```{r}
library(ggplot2)
library(gtable)
library(gridExtra)
library(grid)

library(cowplot)
email_sharing_plots <- 
plot_grid(sharing_by_year, 
                     sharing_by_student,
                     sharing_by_country,
                     sharing_by_journal,  
          axis = "lt",
          ncol = 2, 
          align = "hv", 
          labels = "AUTO")

save_plot("../figures/email_sharing_plots.svg", 
          email_sharing_plots,
          base_height = 8)
```

```{r fig.cap="Summary of results of email requests for data"}
knitr::include_graphics("../figures/email_sharing_plots_hand_edit.png")
```

@@

```{r}
jas_article_survey <- readr::read_csv("../data/raw_data/jas_article_survey.csv")

@@

```{r}
# correlation of data type and location of sharing
file_type_location <- 
jas_article_survey %>% 
  mutate(file = strsplit(as.character(`What file format?`), ",")) %>% 
  unnest(file) %>% 
  filter(file != "na" & `Where is the data?` != "na") %>% 
  select(file,  `Where is the data?`) %>% 
  mutate(location = case_when(
    grepl("sup", tolower(`Where is the data?`)) ~ "Supplementary",
    grepl("tab", tolower(`Where is the data?`)) ~ "Table",
    grepl("www|dryad|osf", tolower(`Where is the data?`)) ~ "Repository",
  )) %>% 
  mutate(file = case_when(
    grepl("DOCX|doc", file) ~ "DOCX",
    grepl("XLSX| XLSX", file) ~ "XLSX",
    TRUE ~ file
    ))

plot_file_type_location <- 
file_type_location %>% 
  group_by(file, location) %>% 
  tally() %>% 
  ggplot(aes(file, location)) + 
  geom_point(aes(size = n),
             colour = viridis(2)[2],
             shape = 16) +
  theme_bw() + 
  xlab("") + 
  ylab("") +
  scale_size_continuous(range=c(7,25)) + 
  geom_text(aes(label = n)) +
  theme(legend.position = "none") 
```

```{r}
# correlation of data type and location of sharing
data_type_location <- 
jas_article_survey %>% 
  filter(`data type?` != "na" & `Where is the data?` != "na") %>% 
  select(`data type?`, `Where is the data?`) %>% 
  mutate(data_type = case_when(
    grepl("comp", tolower(`data type?`)) ~ "Compositional",
    grepl("c14|osl", tolower(`data type?`)) ~ "Dating",
    grepl("dna", tolower(`data type?`)) ~ "DNA",
    TRUE ~ `data type?`
  )) %>% 
   mutate(location = case_when(
    grepl("sup", tolower(`Where is the data?`)) ~ "Supplementary",
    grepl("tab", tolower(`Where is the data?`)) ~ "Table",
    grepl("www|dryad|osf", tolower(`Where is the data?`)) ~ "Repository"
  ))

plot_data_type_location <- 
data_type_location %>% 
  group_by(data_type, location) %>% 
  tally() %>% 
  filter(n >= 2) %>% 
  ggplot(aes(data_type, location)) + 
  geom_point(aes(size = n),
             colour = viridis(2)[1],
             shape = 16) +
  theme_bw() + 
  xlab("") + 
  ylab("") +
  scale_size_continuous(range=c(7,25)) + 
  geom_text(aes(label = n), colour = "white") +
  theme(legend.position = "none") 
```

```{r  fig.cap="Summary of relationships between file types, data types, and data locations"}
plot_data_type <- plot_grid(plot_file_type_location, 
                            plot_data_type_location,
                      labels="AUTO", ncol = 1)
plot_data_type
```

@@

A full treatment of the ethical issues of data sharing is beyond the scope of this paper, but we can highlight some points specifically relevant to the question of when _not_ to share archaeological data. Much of the discussion on ethics and data sharing generally relates to appropriate attribution (assigning credit for the data) in data reuse (Duke et al. 2013). The default position for archaeological data, like data from many other sciences, is that they are part of the global human patrimony, and should be accessible to other scholars and the public (Zimmerman et al. 2003; Vitelli & Colwell-Chanthaphonh 2006). However, archaeologists, like other researchers working with local and indigenous communities, governments, and international collaborators, generate research data that is intellectual property with many stakeholders. With diverse stakeholders, archaeological data can become entangled in economics, nationalism, cultural politics, and identity (Nicholas and Wylie 2009). Often, one of these stakeholders is an indigenous community, and archaeologists may find differences between Western and indigenous conceptions of what data are, how they may be generated, used, shared, and 'owned' (Harding et al. 2012).

One of these differences is that Western researchers often have individual autonomy in sharing data, and once data are shared, it is free for all to use. In contrast, within indigenous communities, data are part of the group's identity and property, akin to tangible hertiage (Nicholas 2014), with certain members entrusted to keep that data on behalf of the group, and sometimes it may be inappropriate for these people to share the data with other members of the group, or beyond the group (Tsosie 2007). Differences between Western and indigenous concepts or data are not only cultural - substantial inequalities exist in access to resources for converting archaeological data into economic and political benefit (e.g. graduate degrees, employment, publications, grants, promotions). On one hand, making archaeological data openly available benefits indigenous communities, because it removes some of the financial and technical barriers to accessing that data (e.g. subscription fees to journals). On the other hand, this may be little more than a useless courtesy (Hymes 1972), because simply making data openly available does not give indigenous people access to the contexts where the data can be used to generate economic and political benefits. Thus, open data has different values for Western and indigenous communities, and if archaeologists want to make a compelling case to their indigenous partners for data sharing, they should also make a committement to reducing some of the inequalities in making use of the data, for example, by providing some training in data analysis, or adapting the data into a context that is valued by the indigenous community.

In short, there is more to the ethics of data sharing than avoiding risk of damage. Redacting the locations of archaeological sites to prevent looting as an obvious case of preparing data for sharing to reduce the risk of damage to the sites. However, because of cultural, economic and political differences between Western archaeologists and the communities they work with, an archaeologist cannot consider their own risk assessment to be comprehensive and exhaustive (Nicholas and Wylie 2012). It is crucial for archaeologists to discuss data sharing plans during negotiatations with representatives authorized by the indigenous peoples whose cultural heritage is the subject of investigation. This negotiation process is important because only indigenous peoples themselves can identify potential adverse outcomes to data sharing, and they can do this only if they understand the proposed research and anticipated results. The concrete outcome of this process is often a Memorandum of Understanding between the indigenous community and the archaeologist's employer that documents the ourcome of negotiations about data sharing. If the archaeologist's indigenous partners have reasons not to share data in public, then the archaeologist should respect that position, and note in publication that the data are restricted at the request of the indigenous collaborators.

@@

Hart, E. M., Barmby, P., LeBauer, D., Michonneau, F., Mount, S., Mulrooney, P., Hollister, J. W. (2016). Ten Simple Rules for Digital Data Storage. PLOS Computational Biology, 12(10), e1005097. https://doi.org/10.1371/journal.pcbi.1005097
